---
title: "NRR - Art. 8 Urban Tree Canopy Cover calculation using Woody Vegetation Layer"
author:
    - name: "Karl Ruf"
      email: "ruf@space4environment.com"
    - name: "Luis Cadahía"
      email: "Luis.CADAHIA-LORENZO@ec.europa.eu"
output:
  html_document:
    df_print: paged
include-in-header: true
affiliation: "European Topic Centre - Data integration and digitalisation (ETC-DI) | Joint Research Centre (JRC)"
date: "2026-06-08"
license: "CC-BY-4.0"
---

# Overview

This notebook presents a follow through example for the calculation of the Nature Restoration Regulation Article 8 Tree Canopy Cover (TCC) indicator using the designated Copernicus alternative - the **Woody Vegetation Layer (WVL)**.

The example shows how to calculate the TCC indicator using the WVL layer for two possible Urban Ecosystem Area (UEA) definitions using (a) Local Administrative Units (LAU) classified as "Cities" as well as "Towns and suburbs" and (b) Urban centres and Urban clusters inside LAUs classified as "Cities as well as "Towns and suburbs" for the example country of Luxembourg. 

Based on the regulation both calculations thereby refer to the maximum (a) and minimum (b) potentially possible reference area designation.

## General workflow

The calculation process features the following sequential steps:

1. Merging the urban centres and urban clusters.
2. Preparing the LAU vector layer, including the following steps:
  + Filtering the LAU by country 
  + Joining the degree of urbanisation and extracting classes cities (1) as well as towns and suburbs (2).
  + Computing area of urban centres and clusters within the LAU. 
3. Masking the WVL.
4. Computing the TCC for (a) the entire LAU and (b) the urban centres and clusters within the LAU.
5. Exporting the TCC data to Reportnet3 compliant format.

## Input data

For calculation of the indicator we require 5 datasets

| Dataset Name | Datatype | Scale / Resolution | Description |
|--------------|----------|-------------------|-------------|
| [Local Administrative Units](https://ec.europa.eu/eurostat/web/gisco/geodata/statistical-units/local-administrative-units) | Vector | 1:1M | Geospatial Local Administrative Units (LAU) to be combined with the 2024 classified EUROSTAT’s degree of urbanisation typology [LAU DEGURBA tabular list](https://ec.europa.eu/eurostat/web/nuts/local-administrative-units). Note this is the public version of the geospatial LAU at scale 1:1M; the internal 1:100,000 version can be requested by country representatives from Eurostat directly or downloaded from the [NRP Reportnet 3 dataflow](https://reportnet.europa.eu/dataflow/1906) in the 'Dataflow help' section. |
| [Urban centres](https://ec.europa.eu/eurostat/web/gisco/geodata/population-distribution/clusters) | Raster | 1 km | Urban centres and urban clusters are groups of 1 km² population grid cells that share similar characteristics, based on a combination of their population density and geographical contiguity. |
| [Urban clusters](https://ec.europa.eu/eurostat/web/gisco/geodata/population-distribution/clusters) | Raster | 1 km | Urban centres and urban clusters are groups of 1 km² population grid cells that share similar characteristics, based on a combination of their population density and geographical contiguity. |
| [Woody Vegetation Layer](https://land.copernicus.eu/en/products/high-resolution-layer-small-landscape-features/woody-vegetation-layer-2021) | Raster | 5 m | Binary mapping of woody vegetation of any type without any differentiation of height, size, or land use. Note the 2021 version may be used for draft NRP. |


The WVL can be downloaded as pre-packaged collections for countries or alternatively via an API. For download you will need to authenticate yourself. CLMS Website authentication is handled using EU Login, European Commission’s user authentication service. Please see https://ecas.ec.europa.eu/cas/help.html for further information.

We here use the WVL version with EU coverage to accommodate the option that this script can be applied universally to all Member States.



```{r setup, include=FALSE}

# Install package manager if needed
if (!requireNamespace("pacman", quietly = TRUE)) {
  install.packages("pacman")
}

# Load required libraries
pacman::p_load(here, sf, tidyverse, terra, tidyterra, exactextractr, readxl, writexl, htmltools, leaflet)

# Hide terra´s progressing bar
terraOptions(progress = 0)


# Country boundary extent based on Reportnet3 Raster templates

eu_cntr_bb <- data.frame(
  cntr_code = c(
    "AT","BE","BG","CY","CZ","DE","DK","EE","ES","ES","FI","FR",
    "FR","FR","FR","FR","FR","GR","HR","HU","IE","IT","LT","LU",
    "LV","MT","NL","PL","PT","PT","PT","RO","SE","SI","SK"
  ),
  file_name = c(
    "Art8_RasterTemplate_10m_AT.tif",
    "Art8_RasterTemplate_10m_BE.tif",
    "Art8_RasterTemplate_10m_BG.tif",
    "Art8_RasterTemplate_10m_CY.tif",
    "Art8_RasterTemplate_10m_CZ.tif",
    "Art8_RasterTemplate_10m_DE.tif",
    "Art8_RasterTemplate_10m_DK.tif",
    "Art8_RasterTemplate_10m_EE.tif",
    "Art8_RasterTemplate_10m_ES.tif",
    "Art8_RasterTemplate_10m_ES_CA.tif",
    "Art8_RasterTemplate_10m_FI.tif",
    "Art8_RasterTemplate_10m_FR.tif",
    "Art8_RasterTemplate_10m_FR_GF.tif",
    "Art8_RasterTemplate_10m_FR_GP.tif",
    "Art8_RasterTemplate_10m_FR_MQ.tif",
    "Art8_RasterTemplate_10m_FR_RE.tif",
    "Art8_RasterTemplate_10m_FR_YT.tif",
    "Art8_RasterTemplate_10m_GR.tif",
    "Art8_RasterTemplate_10m_HR.tif",
    "Art8_RasterTemplate_10m_HU.tif",
    "Art8_RasterTemplate_10m_IE.tif",
    "Art8_RasterTemplate_10m_IT.tif",
    "Art8_RasterTemplate_10m_LT.tif",
    "Art8_RasterTemplate_10m_LU.tif",
    "Art8_RasterTemplate_10m_LV.tif",
    "Art8_RasterTemplate_10m_MT.tif",
    "Art8_RasterTemplate_10m_NL.tif",
    "Art8_RasterTemplate_10m_PL.tif",
    "Art8_RasterTemplate_10m_PT.tif",
    "Art8_RasterTemplate_10m_PT_AZ.tif",
    "Art8_RasterTemplate_10m_PT_MA.tif",
    "Art8_RasterTemplate_10m_RO.tif",
    "Art8_RasterTemplate_10m_SE.tif",
    "Art8_RasterTemplate_10m_SI.tif",
    "Art8_RasterTemplate_10m_SK.tif"
  ),
  xmin = c(
    4284000,3798000,5311000,6340000,4469000,4030000,4198000,5006000,
    2759000,1546000,4743000,3208000,98000,626000,690000,314000,
    493000,5124000,4571000,4785000,2923000,4054000,5006000,4013000,
    4991000,4700000,3857000,4594000,2634000,942000,1790000,5111000,
    4375000,4571000,4825000
  ),
  xmax = c(
    4856000,4067000,5813000,6529000,4962000,4674000,4652000,5369000,
    3835000,2048000,5405000,4286000,432000,715000,738000,380000,
    534000,5957000,5064000,5281000,3276000,5051000,5381000,4073000,
    5439000,4740000,4137000,5314000,2979000,1330000,1890000,5859000,
    4974000,4830000,5236000
  ),
  ymin = c(
    2594000,2940000,2118000,1592000,2834000,2683000,3495000,3924000,
    1438000,940000,4086000,2025000,233000,1750000,1591000,7633000,
    8554000,1423000,2163000,2548000,3325000,1385000,3506000,2932000,
    3714000,1424000,3077000,2944000,1728000,2249000,1203000,2389000,
    3585000,2482000,2765000
  ),
  ymax = c(
    2893000,3169000,2496000,1764000,3115000,3553000,3852000,4194000,
    2470000,1146000,5309000,3137000,638000,1827000,1647000,7692000,
    8610000,2218000,2627000,2897000,3726000,2668000,3802000,3017000,
    3985000,1459000,3413000,3558000,2300000,2795000,1544000,2937000,
    5138000,2662000,2994000
  ),
  epsg = c(
    3035,3035,3035,3035,3035,3035,3035,3035,
    3035,3035,3035,3035,32622,32620,32620,32740,
    32738,3035,3035,3035,3035,3035,3035,3035,
    3035,3035,3035,3035,3035,3035,3035,3035,
    3035,3035,3035
  ),
  stringsAsFactors = FALSE
)


```


## 1. Merge urban centres and urban clusters

We will be computing our TCC for the urban centres and urban clusters inside the LAUs. For this we have to merge the two individual clusters to one common dataset. Throughout the process you will have to update the paths to the datasets specified in "Input data". Below you can specify your 2-digit country code which will set processing extent window that matches your country to obtain national results.

Both raster datasets contain integer values representing the respective population of the contiguous group of cells.

We can thus use simple conditional logic to combine the two layers. Areas with values above 0 in either layer are combined. 

```{r Combine urban centres and clusters, warning=FALSE}

# We start by setting our country 2-digit code and processing extent used reading only the area that we need.
cc <- 'LU' # Change here. Use EL not GR for Greece.


# Next we will retrieve the bounding box for the country. 
# This has been pre-processed from the templates available on Reportnet3.
# Coordinates were rounded to min/max and dividable by 1000 to match resolution or UC/UC data.
# For FR, ES and PT overseas territories refer to table under setup to identify the correct extent and set 
# by using cntr_bb <- ext(xmin,xmax,ymin, ymax).

cntr_bb <- eu_cntr_bb %>% 
  filter(cntr_code == cc) %>% 
  select(xmin, xmax, ymin, ymax) %>% 
  as.vector() %>% 
  unlist() %>% 
  ext()
  

# Load high density clusters
ucentres <- rast(here("data", "HDENS_CLST_2021.tif"),
                 win = cntr_bb
                 )

# Load medium density clusters
uclusters <- rast(here("data", "URBAN_CLST_2021.tif"),
                  win = cntr_bb
                  )


# Merge with using conditional logic
ucuc <- terra::ifel(ucentres > 0 | uclusters > 0, 1, NA, 
                        datatype = "INT1U")


```


```{r UCUC leaflet, warning=FALSE, message=FALSE, echo=FALSE}

# Set to NA as to avoid being displayed as background value
ucentres[ucentres == 0] <- NA
uclusters[uclusters == 0] <- NA

leaflet::leaflet() %>% 
  addProviderTiles(
    providers$CartoDB.Positron,
    group = "Basemap"
  ) %>%
  
  addMapPane("clusters", zIndex = 410) %>%
  addMapPane("centres", zIndex = 420) %>%
  
  addRasterImage(
    ucuc,
    colors = c("#f03b20", "#0000FF"),
    opacity = 1,
    project = TRUE,
    group = "Combined UCUC"
  ) %>% 
  
    addRasterImage(
    uclusters,
    colors = c( "#feb24c"),
    opacity = 1,
    project = TRUE,
    group = "Urban Clusters",
    options = leafletOptions(pane = "clusters")
  ) %>% 

  addRasterImage(
    ucentres,
    colors = c( "#e31a1c"),
    opacity = 1,
    project = TRUE,
    group = "Urban Centres",
    options = leafletOptions(pane = "centres")
  ) %>% 
  
  addLayersControl(
    overlayGroups = c(
      "Urban Centres",
      "Urban Clusters",
      "Combined UCUC"
    ),
    options = layersControlOptions(collapsed = FALSE)
  ) %>% 
  hideGroup(c("Combined UCUC", "Urban Clusters"))


```


## 2. Prepare the Local Administrative Units 

We first start with preparing the LAU by filtering those which are either cities (1) or towns and suburbs (2). 

It is up to the countries to define whether 

a. the full LAU area is defined as Urban Ecosystem Area (UEA) or,
b. only the area of urban centres and clusters within the LAU shall be defined as UEA.

The UEA is the denominator in the calculation of the tree canopy cover. In addition to these two options covered here, countries may also customise the UEA definition by merging neighbouring LAUs or adding a buffer. 

For option a) the UEA is defined directly by the area of the polygon geometry while for b) we need to calculate the urban centres and cluster area within the LAU.

For the latter we use the exact_extract function from the "exactextractr" package (Baston 2023).

This library allows to calculate the coverage fraction, accounting for fractional pixel cover and is thus the most accurate tool for UEA area calculation.


```{r LAU preprocessing, warning=FALSE, message=FALSE}

# Create the extraction query for the LAU file
cntr_query <- paste0(
    "SELECT * FROM 'LAU_RG_01M_2024_3035.gpkg' WHERE CNTR_CODE = '",
    cc,
    "'"
  )

# Load the LAUs geometry layer - We can use a query upon load to filter only the national LAUs
lau_geom <- st_read(
                here("data", "LAU_RG_01M_2024_3035.gpkg"),
                quiet = TRUE,
                query = cntr_query,
                fid_column_name = "fid"
                )

# Load the respective LAU classification
degurba <- read_xlsx(
  here("data", "EU-27-LAU-2024-NUTS-2024.xlsx"), 
  sheet = cc
  ) 


# We join the Degree of Urbanisation classification to the geometries...
lau_degurba <- left_join(lau_geom, degurba, by = c("GISCO_ID" = "EU LAU CODE")) 


#...extract only cities (1) and towns and suburbs (2) and do some attribute cleaning
lau_degurba_1_2 <- lau_degurba %>% 
  filter(DEGURBA < 3) %>% 
  rename(uea_lau_km2 = AREA_KM2) %>% 
  select(
    
    fid,
    GISCO_ID,
    CNTR_CODE,
    LAU_NAME,
    DEGURBA,
    uea_lau_km2
    
  )


# Then we compute the Urban Ecosystem Area (UEA) based on the urban centres and clusters within the LAUs (Option b described above)
uea_ucuc <- exact_extract(ucuc,
                                 lau_degurba_1_2,
                                 fun         = "count",
                                 append_cols = c("GISCO_ID"),
                                 progress = FALSE)

# The pixel count reflects the coverage fraction. In addition, because we used the 1x1km ucuc layer, pixel size equals 1 km² and, hence, the count equals the area in km².

uea_ucuc <- uea_ucuc %>% rename(uea_ucuc_km2 = "count")
 
# Join the computed area back to the LAU vector layer
lau_degurba_1_2 <- left_join(lau_degurba_1_2,
                                 uea_ucuc,
                                 by = "GISCO_ID")



```


## 3. Masking of WVL data

Now that we have our combined urban centres and urban clusters layer prepared we want to use this to mask the WVL. Therefore we first need to align the resolution of both layers. We thus resample the 1km UCUC from 1km to 5m resolution.

```{r Mask WVL, warning=FALSE, message=FALSE}

# Load the WVL 
wvl <- rast(here("data", "WVL_S2021_005m_eu_03035_V01_R03.tif"),
            win = cntr_bb)


# Next we resample the urban centres and urban clusters to the 5m resolution.
# Note that you may also slice pixels using terra::disagg 
# with a disaggregation factor of 200 (1000m / 5m = 200)
# However this may entail additional cropping/extending steps to align with the ucuc mask.


ucuc_5m <- terra::resample(ucuc,
                           wvl,
                           method = "near")

# The disaggregated 5m raster can now be used as a mask
wvl_ucuc <- mask(wvl,
                 ucuc_5m)


```


## 4. Tree Canopy Cover (TCC) calculation with WVL

The final tree canopy cover is calculated using exact_extract function from the exactextractr package (Baston 2023).

This library allows for fractional pixel counting and is thus the most accurate tool for quantification of WVL pixels.

Statistics can be viewed as table and map below.

```{r Tree Canopy Cover quantification, warning=FALSE, message=FALSE}

# Zonal statistics for full LAU extent
tcc_lau_df <- exact_extract(wvl,
                                 lau_degurba_1_2,
                                 fun         = "mean", # For a binary raster, the mean equals the fraction covered.
                                 append_cols = c("fid",
                                                 "GISCO_ID", 
                                                 "CNTR_CODE", 
                                                 "LAU_NAME",
                                                 "DEGURBA", 
                                                 "uea_lau_km2",
                                                 "uea_ucuc_km2"),
                                 progress = FALSE)


tcc_lau_df <- tcc_lau_df %>% 
  mutate(tcc_lau_pct = mean*100,
         tcc_lau_km2 = uea_lau_km2*mean) %>% 
  select(GISCO_ID, tcc_lau_pct,tcc_lau_km2)



# Zonal statistics for urban centres and urban clusters only
tcc_ucuc_df <- exact_extract(wvl_ucuc,
                                 lau_degurba_1_2,
                                 fun         = "mean", # For a binary raster, the mean equals the fraction covered.
                                 append_cols = c("fid",
                                                 "GISCO_ID", 
                                                 "CNTR_CODE", 
                                                 "LAU_NAME",
                                                 "DEGURBA", 
                                                 "uea_lau_km2",
                                                 "uea_ucuc_km2"),
                                 progress = FALSE)


tcc_ucuc_df <- tcc_ucuc_df %>% 
  mutate(tcc_ucuc_pct = mean*100,
         tcc_ucuc_km2 = uea_ucuc_km2*mean) %>% 
  select(GISCO_ID, tcc_ucuc_pct, tcc_ucuc_km2)

# Combine the two computations 
tcc_stats <- left_join(tcc_lau_df,  tcc_ucuc_df, by = "GISCO_ID")

# Join to the LAU vector layer
tcc <- left_join(lau_degurba_1_2,  tcc_stats, by = "GISCO_ID")

# Show the final results
tcc %>% st_drop_geometry() %>%  
  select(GISCO_ID, LAU_NAME, DEGURBA, uea_lau_km2, uea_ucuc_km2, tcc_lau_pct, tcc_ucuc_pct, tcc_ucuc_km2, tcc_ucuc_km2)
```


```{r Tree Canopy Cover leaflet, warning=FALSE, message=FALSE, echo=FALSE}

# Reproject for display in leaflet
tcc_epsg4326 <- st_transform(tcc, "EPSG:4326" )

# Set colouring
pal_dgurba <- colorFactor(
  palette = c("red", "yellow"),
  domain = c(1, 2)
)


# Add a leaflet for display
leaflet(tcc_epsg4326) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addRasterImage(
    ucuc_5m,
    colors = c("#f03b20", "#feb24c"),
    opacity = 1,
    project = TRUE,
    group = "UCUC"
  ) %>% 
  addRasterImage(
    wvl_ucuc,
    colors = c("transparent", "#75dd00"),
    opacity = 1,
    project = TRUE,
    group = "WVL - UCUC"
  ) %>% 

  addPolygons(
    fillColor = ~pal_dgurba(DEGURBA),
    fillOpacity = 0.7,
    color = "#444444",
    weight = 1,
    group = "LAU",
    label = ~lapply(
      paste0(
        "<b>LAU name:</b> ", tcc_epsg4326$LAU_NAME, "<br>",
        "<b>LAU area:</b> ", round(uea_lau_km2, 2), " km²<br>",
        "<b>UCUC area:</b> ", round(uea_ucuc_km2, 2), " km²<br>",
        "<b>TCC/WVL LAU:</b> ", round(tcc_lau_pct, 1), "%<br>",
        "<b>TCC/WVL UCUC:</b> ", round(tcc_ucuc_pct, 1), "%"
      ),
      htmltools::HTML
    ),
    highlightOptions = highlightOptions(
      weight = 3,
      color = "black",
      bringToFront = TRUE
    )
  ) %>%
  addLegend(
    position = "bottomright",
    colors = c("red", "yellow"),
    labels = c(
                "1 - Urban Centre",
                "2 - Urban Cluster"
              ),
    title = "Degree of Urbanisation",
    opacity = 1
  ) %>% 
    leaflet::addLayersControl(
    overlayGroups = c(
      "LAU",
      "UCUC",
      "WVL - UCUC"
    ),
    options = layersControlOptions(collapsed = FALSE)
  ) %>% 
  leaflet::hideGroup(c("WVL - UCUC", "UCUC"))


```

## 5. Exporting to Reportnet3 compliant format

As a final step we need to perform a small chain of file format operations to match the Reportnet3 delivery format.
For this we need to specify which calculation method was selected (see a or b in section 2) and the destination of the exported file.
The "lau_id" attribute is generated from the GISCO ID contained in the LAU polygon layer.

Some additional attributes are added for compliance purposes, but are not filled in this example (uea_id and area_ugs).
The fid is in this example inherited from the source LAU polygon layer and is not recalculated during export of the file. 

```{r File formating and Data export, warning=FALSE, message=FALSE}

# Generate t8_UEAPolygons ready for upload to Reportnet 3 NRR dataflow

uea_definition <- "ucuc" # Enter either "ucuc" or "lau"  See cases a or b in section 2

fn <- here("f03_products", "NRR_Art8",  "t8_UEAPolygons.gpkg") # Change name / path here


# Change file format to NRR delivery structure
t8_UEAPolygons <- tcc %>% 
    mutate(
           uea_id = NA_character_, # Empty for now
           lau_id = paste("lau2024", 
                          str_split_i(GISCO_ID, "_", 1), 
                          str_split_i(GISCO_ID, "_", 2),
                          sep = "."),
           area_tcc = if (uea_definition == "ucuc")
                          round(tcc_ucuc_km2, 5)
                      else
                          round(tcc_lau_km2, 5),
           area_ugs = NA_real_ # Empty for now attribute specifying the Urban Green Space not covered here
          ) %>%  
    rename(geometry = geom) %>% 
    select(fid, geometry, uea_id, lau_id, area_tcc, area_ugs)
  
   
  # Set native sf "geom" geometry encoding name to "geometry"
  st_geometry(t8_UEAPolygons) <- "geometry"

    
  # Export
  st_write(t8_UEAPolygons, 
           fn, 
           fid_column_name = "fid",
           quite=TRUE
           )
  

```

