---
title: "R Notebook"
output: html_notebook
author: "Code: Luis Cadahia, Refinement: Karl Ruf"
---

# NRR - Art. 8 Urban Tree Canopy Cover calculation using Woody Vegetation Layer

This notebook presents a follow - through example for the calculation of the Nature Restoration Regulation Article.8 Tree Canopy Cover (TCC) indicator. Using the designated Copernicus alternative - the **Woody Vegetation Layer (WVL)**.

The example shall cover a full follow through example on how to to calculate the TCC indicator using the WVL layer for Urban centres and Urban clusters inside Local Administrative Units (LAU) for the example of Luxembourg.


For calculation of the indicator we require 4 datasets

| Dataset Name | Datatype | Scale / Resolution | Description |
|--------------|----------|-------------------|-------------|
| [Local Administrative Units](https://ec.europa.eu/eurostat/web/gisco/geodata/population-distribution/degree-urbanisation) | Vector | 1:1M | Administrative units classified according to EUROSTAT's degree of urbanisation typology. Note this is the public version at scale 1:1M; the internal 1:100,000 version can be requested by country representatives from Eurostat directly. |
| [Urban centres](https://ec.europa.eu/eurostat/web/gisco/geodata/population-distribution/clusters) | Raster | 1 km | Urban centres and urban clusters are groups of 1 km² population grid cells that share similar characteristics, based on a combination of their population density and geographical contiguity. |
| [Urban clusters](https://ec.europa.eu/eurostat/web/gisco/geodata/population-distribution/clusters) | Raster | 1 km | Urban centres and urban clusters are groups of 1 km² population grid cells that share similar characteristics, based on a combination of their population density and geographical contiguity. |
| [Woody Vegetation Layer](https://land.copernicus.eu/en/products/high-resolution-layer-small-landscape-features/woody-vegetation-layer-2021) | Raster | 5 m | Binary mapping of woody vegetation of any type without any differentiation of height, size, or land use. |


The WVL can be downloaded as pre-packaged collections for countries or alternatively via an API.For download you will need to authenticate yourself. CLMS Website authentification is handled using EU Login, European Commission’s user authentication service. Please see https://ecas.ec.europa.eu/cas/help.html for further information.


```{r setup}

# Install package manager if needed
if (!requireNamespace("pacman", quietly = TRUE)) {
  install.packages("pacman")
}

# Load required libraries
pacman::p_load(here, sf, tidyverse, terra, tidyterra, exactextractr, skimr, extrafont, scales, writexl, httr2,jsonlite)

# Folders
output_folder <- here('f02_data', 'f03_output')
tcd_folder <-  here('f02_data','f01_input', 'TCD', 'TCD2024')


# Degree of urbanisation Level 1 / 2
lau <- st_read(here('f02_data', 'f02_processing', 'T20_db.gpkg'), layer='tcd24_wvl_comparison__dgurba_rg_01m_2021_3035_selection')

dgurba <- rast(here('f02_data', 'f01_input', 'URBAN_CLST_2021.tif'))





# Reference  Data 
wvl <- rast("R:/INSPIRE_Annex_II/02_Land_cover/08_Copernicus_High_Resolution_Layer/01_Forest/small_woody_features/2021/26691/Results/WVL_2021_005m_eu_03035_V01_R01/WVL_2021_005m_eu_03035_V01_R01/WVL_2021_005m_eu_03035_V01_R01.tif")

# TCD 21 / 24 loaded later as vrt

# EEA 100km grid
eeagrid <- https://sdi.eea.europa.eu/data/a373b065-66d8-45ae-86af-6618cc6792cb


# Test cities: Coblence(DE), Mainz (DE), Linz (AT), Reggio d'Emilia (IT), Gdànsk (PL)
lau_id_selection <- c('7315000', '7111000', '40101', '35033', '1004221436101' ) #Note LAU´s stored as string


```

```{r Combine urban centres and clusters}
ucentres <- rast(here("f02_data", "f01_input",  "HDENS-CLST-2021", "HDENS_CLST_2021.tif"))
ucentres

uclusters <- rast(here("f02_data", "f01_input",  "URBAN-CLST-2021", "URBAN_CLST_2021.tif"))
uclusters


# Check extents and match if necessary ----

ext(ucentres)
ext(uclusters)
crs(ucentres) == crs(uclusters)
res(ucentres) == res(uclusters)

# The extent of 'uclusters' is larger than ucentres, so we need to go for the biggest extent, to keep everything

combined_ext <- ext(terra::union(ext(ucentres), ext(uclusters)))

# Extend ucentres to the larger extent

ucentres <- extend(ucentres, combined_ext)

# Check again

ext(ucentres) == ext(uclusters)
crs(ucentres) == crs(uclusters)
res(ucentres) == res(uclusters)

# All TRUE, so we proceed to merging them.


# Merge with terra::ifel ----
# The original rasters don't have 0 and 1, for some reason, each centre or cluster has either a seemingly random number or zero. That number indicates that there is cluster or centre there, so for the merge with ifelse, I used > 0 instead of == 1.

tic()

UCUC_1km <- terra::ifel(ucentres > 0 | uclusters > 0, 1, NA, 
                        filename = here("f02_data", "f02_processing", "UCUC_1km.tif"),
                        overwrite = TRUE,
                        datatype = "INT1U",
                        progress = TRUE)

toc() # 1.002 sec elapsed


UCUC_1km
names(UCUC_1km) <- "UCUC_1km"
varnames(UCUC_1km) <- "UCUC_1km"
UCUC_1km
```



```{r Data Download}


url <- paste0(
  "https://land.copernicus.eu/api/en/products/high-resolution-layer-small-landscape-features/woody-vegetation-layer-2021?",
  "portal_type=DataSet&",
  "SearchableText=Woody%20Vegetation%20Layer%202021&",
  "metadata_fields=UID&",
  "metadata_fields=downloadable_files&",
  "metadata_fields=Countries"
)

s <- request(url) |>
  req_perform()




{"client_id":"7c929f08feae2f3421fb7bc7af0c0dfb","ip_range":null,"issued":"2026-06-02T13:22:18.809326","key_id":"faaee6e7bcfad1a8e67c7abf3695c931febf7013","private_key":"-----BEGIN RSA PRIVATE KEY-----\nMIIEpAIBAAKCAQEA45rHh1MW/95Nmv7OxWzpmcnSmruxxoG79TXVq7SUVWVG1eBL\n83Zlp25jOBSIbafGKUTiUPiOsDmjwRzklYq6xx/4NoPbciMvpizOaKYUV7XnBE9J\n6MhHR77Sr9chmXFu4CFN5iV8KgGlIKoXonGaJOPGAkIfX9+KXN+8L6JglJj6sOjj\n135rswh6lk+aX/AlZi+ungPpJu5Kw2RLct6ATz0dA8h1ojPjCAfNdVq/1E8bptkQ\nj/HZEFv7w/g3GnKzurkJQPdPVntNXWJPEDpfAhGGs3n3RbopEnaRl5pavTLWZSN7\nYsOkGi8YkZC5F4wXV9ETW7hJ2wDrVikWJS8pxQIDAQABAoIBAFed0zZ8XWu1G4HR\nmi503T4266xmGar1cTIVfcICG2H+QTT/Ca3Yw4wfGfHSMayou8dSPZ72+BLNtYSX\nB0Wr6HfUDjyKKNV4xxR8xh6R5F3ZZ+uITEbgplPvmPJY6hXEDW8ZUwAIKRKQnI0E\n/WEqwzB1QFnJM4O4LW7QMATUSMKaoWW89xHI4vN+EfqRxPZXtraFE9x3O8Rb+MPr\n4LvvfafhqXDEyuVnR4nw/6tWSW5oX5xLib8yQSKIqzxaFZOF6Uju+dwLro/9Vrnz\nw4rlHTKGp9fiRmuAHC4PQVtMxBC8i5YLKi6Cv95mE9iHwPCaOF1KEuQ9UbVRV5mN\nj6PU9UcCgYEA+JNfCmvqNReL+t7Bx668B3HznP3BmlTgpVT0TR7YhEm1ixWvu8iP\n5omTS6+lDkcybrh/MY0+6R/8gZ3qX10tcEVrCRz+Y6LJTUXDKpkbS4bV5TC0SvYv\nfna0+zHyqNzkBAyigVAw4jwqTwvygtbJZskhYrR6vCjcTVpxmjCqe/MCgYEA6mcP\n0gPo9KleF8rgfgw5AJnKClCBdkhBmVEQlmBmlyjApnA4/My8jvejaApgbGfmIPaU\nN6je0h5KkOS87zBE/x434v9Uh24bQCTUEhEVhs7+s/p6Zc4G4dktnbKOBg1/YMlx\n/fzUEkHjdgyumZx8KA0kWegDZ7qA/QieCbqFSWcCgYEAuqM0Tgfez0RBxvgwOTS6\nW0C08GAbWDx4e8rAr5MEcLdDRw3eZ7lcSnVrkf1sib7tX/LJFbR+0t1AYzXlfvrg\nHhcRGcOaQcc9K/bAZiCVPO0ddrtN86bK3EJW+vByyVUpIt01urAyM/uGrNlEmfaZ\nO8KyfYETUmm/l5jcCTCXIP8CgYABk8n2dtJFezJH5LSYBIsJpv9GDH6ihh/y4SnE\naUencr3BIPYxNUt7hs0RsSgu8pAI3r/mk/tT3KJR0Ug4L9H3CYQEwgUWmDYUu2q1\n6GmFq7TpDyFtt/Ix6N02t3HfNo4J37ElrWLb8JeQPsuZ+wSW3GfHNxl2xDmD+l/d\n3uRwtwKBgQCmTt7l4jvJYrvIxel0JgasT0CHeNCjh0CLjreXTcG0QLwgFNk/2lDr\nSpW2w4rt2eRb+7IXYCHFVfLKHcbePACx5lWcMDJHTRnY4IJgCO4edYAWX/EJXG8l\nG9eqLt0aFGUd0x8KlMMfKIizP+f49lVydyPoiVMtAR3brr+r57c5+w==\n-----END RSA PRIVATE KEY-----\n","title":"WVL_download","token_uri":"https://land.copernicus.eu/@@oauth2-token","user_id":"nrufkarl"}

```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
